
# Style guide

### What is a coding style?
    
``Good coding style is like using correct punctuation. You can manage without
it, but it sure makes things easier to read.'' [Headley Wickham](http://adv-r.had.co.nz/Style.html).

``thus try to make use of it try to follow the style guides you
may thin it is kind of useless in the beginning however you
will see that it is especially when your functions get more complex
very useful to do so i often spent hours to find my own bugs
just because my coding style was horrible keep that in mind'' Reto, 2019.

```{r, echo = FALSE, fig.align = "center"}
options(prompt = " ", continue = " ")
```

### Special characters

* do not use special characters (Umlauge, apostrophes, ...)
* neither for object names, nor comments!
* you may have noticed it: object names have to start with
   a letter, blanks are not allowed

### Language

Preferably in English. Especially if you hand over your code to
someone else.

### Script file names

Should be meaningful and use the `.R` suffix.

**Good:**

* `fit-linear-models.R`
* `utility-functions.R`
* `session02_exerciseA.R`

**Bad:**

* `foo.R`
* `things.R`
* `exercise.R`

### Object names

Should be lowercase and meaningful! Use underscores to separate
words within a name. If possible use nouns for variables
and verbs for functions.

**Good:**

* `day_one`
* `day_1`
* `gender`
* `height`
* `smallest_male`
* `temp_kelvin`

**Bad:**

* `the_first_day_of_month`
* `DayOne`
* `dayone`
* `djm1`
* `foo`
* `T` (existing name!)
* `c` (existing name!)
* `mean` (existing name!)

### Spaces: infix operators

Put spaces around all infix operators (`+`, `/`, `*`,
`&`, `==`, `|`), around the get operator (`<-`) and `=`' s.

**Well written code:**
```{r, echo = FALSE, results = "hide", fig.align = "center"}
temp <- 5
```

```{r, fig.align = "center"}
temp <- temp * 9 / 5 + 32
volume <- 4 / 3 * pi**3
if (temp == 0) print("cool")
```

**Bad code:**
```{r, fig.align = "center"}
temp<-temp*9/5+32
volume<-4/3*pi**3
if( temp==0) print("cool")
```

### Spaces: colons

One exception: There is no space around colons (`:`, `::`, `:::`):

**Good:**
```{r, fig.align = "center"}
x <- 1:10
```

**Bad:**
```{r, fig.align = "center"}
x <- 1 : 10
```

### Spaces: multiple spaces

Ok to use multiple spaces if it increases readability. E.g.:

**More well-structured ...**
```{r, echo = FALSE, results = "hide", fig.align = "center"}
l <- b <- h <- 5
```

```{r, echo = TRUE, results = "hide", fig.show = "hide", fig.align = "center"}
list(
  total    = l + b + h,
  mean     = (l + b + h) / 3,
  variance = var(c(l + b + h))
)
```

**... than this:**
```{r, echo = TRUE, results = "hide", fig.show = "hide", fig.align = "center"}
list(
  total = l + b + h,
  mean = (l + b + h) / 3,
  variance = var(c(l + b + h))
)
```

### Spaces: brackets

A space before opening round brackets (`(`) and
after closing round brackets (`)`)
_except for function calls_.

**Good:**
```{r, echo = FALSE, results = "hide", fig.show = "hide", fig.align = "center"}
a <- 10
debug <- FALSE
x <- y <- 10
```

```{r, echo = TRUE, results = "hide", fig.show = "hide", fig.align = "center"}
if (a == 4)
if (debug) print(x)
mean(x)
plot(x, y, col = "red")
```

**Bad:**
```{r, echo = TRUE, results = "hide", fig.show = "hide", fig.align = "center"}
if(a==4)
if(debug)print(x)
mean (x)
plot (x,y,col="red")
```

### Spaces: commas

Always use spaces after the comma.

**Good:**
```{r, echo = FALSE, results = "hide", fig.align = "center"}
mat <- matrix(1:9, ncol = 3)
y <- rnorm(3)
```

```{r, echo = TRUE, results = "hide", fig.show = "hide", fig.align = "center"}
x <- c(1, 2, 4)
x <- mat[1, ]
plot(x, y, col = "red")
```

**Bad:**
```{r, echo = FALSE, results = "hide", fig.align = "center"}
mat <- matrix(1:9, ncol = 3)
y <- rnorm(3)
```

```{r, echo = TRUE, results = "hide", fig.show = "hide", fig.align = "center"}
x <- c(1,2,4)
x <- mat[1,]
plot(x,y,col = "red")
```

### Opening/closing brackets

Opening curly brackets should never go in a new line,
but should always be followed by a new line.
Final closing curly brackets should be in a separate line.

**Good:**
```{r, echo = FALSE, results = "hide", fig.align = "center"}
a <- 10
```

```{r, echo = TRUE, results = "hide", fig.align = "center"}
if (a == 10) {
  res <- TRUE
} else {
  res <- FALSE
}

myfun <- function(x) {
  # Instructions
}
```

**Bad:**
```{r, echo = TRUE, results = "hide", fig.align = "center"}
if (a == 10)
{
  res <- TRUE
} else
{
  res <- FALSE}
```

### Indention

Use block-wise indention (RStudio supports you).
Suggested by Headley: use two spaces as indent.

**Good:**
```{r, echo = TRUE, results = "hide", fig.align = "center"}
powfun <- function(x) {
  res <- x**y
  return(res)
}
```

**Bad:**
```{r, echo = TRUE, results = "hide", fig.align = "center"}
powfun <- function(x) {
res <- x**y
return(res)
}
```

**For very short statements:**
```{r, echo = FALSE, results = "hide", fig.align = "center"}
a <- 10
```

```{r, echo = TRUE, results = "hide", fig.align = "center"}
powfun <- function(x, y) { return(x^y) }
if (a == 10) { res <- TRUE }
```

```{r, echo = TRUE, results = "hide", fig.align = "center"}
powfun <- function(x, y) return(x^y)
if (a == 10) res <- TRUE
```

### Explicit return

My personal suggestion: explicitly use the \code{return} command
when returning results from a function!

**Good:**
```{r, echo = TRUE, results = "hide", fig.align = "center"}
powfun <- function(x) {
    res <- x^2
    return(res)
}
powfun <- function(x, y) return(x^y)
```

**Bad:**
```{r, echo = TRUE, results = "hide", fig.align = "center"}
powfun <- function(x) {
    res <- x^2
    res
}
powfun <- function(x, y) x^y
```

### Line breaks

Line breaks: lines should not be longer than 80 characters.
Use line breaks for long statements.

**Good:**
```{r, echo = TRUE, fig.align = "center"}
data <- c(firstname  = "Reto",      lastname = "Stauffer",
          city       = "Innsbruck", zipcode  = "6020",
          department = "Department of Statistics")
```

**Bad:**
```{r, echo = TRUE, fig.align = "center"}
data <- c(firstname = "Reto", lastname = "Stauffer", city = "Innsbruck", zipcode = "6020", department = "Department of Statistics")
```

### Nested indention

Use block-wise indention (RStudio supports you).
Use two spaces as indent.

**Good:**
```{r, echo = TRUE, fig.align = "center"}
demo_function <- function(x) {
  if (a == 3) {
    res <- "This"
  } else {
    for (i in 1:10) {
      res <- a + i
    }
  }
  return(res)
}
```

**Bad:**
```{r, echo = TRUE, fig.align = "center"}
demo_function <- function(x) {
if (a == 3) {
res <- "This"
} else {
for (i in 1:10) {
res <- a + i
}
}
return(res)
}
```

### Object assignment

Use `<-` and not `=`.
**Good:**
```{r, echo = TRUE, fig.align = "center"}
a <- 5
powfun <- function(x, y) return(x^y)
```

**Bad:**
```{r, echo = TRUE, fig.align = "center"}
a = 5
powfun = function(x, y) return(x^y)
```

### Comments

* Start to write comments!
* Again: code is more often read then written,
* comments help to understand code (also your own!).
* Use `# ------` or `# ========` to separate code blocks.

```{r, results = "hide", fig.show = "hide", fig.align = "center"}
# -----------------------------
# Define variables
x <- seq(0, 100, length = 100)
y <- sin(x / pi * 180)
# -----------------------------
# Create plot
plot(x, y, col = 5)
```

```{r, echo = FALSE, fig.align = "center"}
options(prompt = "R> ", continue = "+  ")
```

### All together

Both code chunks do the very same. And both will work.
I think there's nothing to say here!

**Good:**
```{r, echo = TRUE, fig.align = "center"}
demo_function <- function(x) {
  if (a == 3) {
    res <- "This"
  } else {
    for (i in 1:10) {
      res <- a + i
    }
  }
  return(res)
}
```

**Very, very bad:**
```{r, echo = TRUE, fig.align = "center"}
DmoFn<-function(x)
{if(a==3) {res <- "This"
}else{
for(i in 1    :10) {
res= a   +i}}
    res}
```


