

# Sort \& Order

Today we will learn about two new functions: **Sort & Order**.

Sometimes it is required to sort an object
(`vector`, `matrix`, `data.frame`, ...).
Not surprisingly the the functions are called `sort` \& `order`.
Note that _R_ base does not provide convenience functions to directly
sort `data.frame`s, however, additional packages like e.g.,
`plyr` or `data.table` have extra functions for it. Nevertheless, we can
also use the base functions to sort our objects without the need to install
and load additional packages.

## Sorting a vector

To bring the values in order we can use `sort`.
From `?sort`/`help("sort")`:

```r
Sort (or order) a vector or factor (partially) into ascending or
descending order. For ordering along more than one variable,
e.g., for sorting data frames, see 'order'.

sort(x, decreasing = FALSE, na.last = NA, ...)
```

As we can see the function `sort` takes 3+ input arguments.
The first one (`x`) is our vector, the following two arguments
are optional and allow to control how the vector will be sorted.
If not specified, the defaults will be used (`decreasing = FALSE`
and `na.last = NA`). We will see how these arguments change the
way the vector is sorted later on.

For demonstration purposes let us create a simple integer
vector with random numbers:
```{r, fig.align = "center"}
# Set randomization seed
set.seed(555)
# Sampling some random integers
x <- sample(1:100, 10)
# Setting element 8 to NA
x[8] <- NA
x
```

Note that I have set the eight element (`x[8]`) to `NA` (missing value)
as we have to take care about missing values when using `sort`.
First, let's simply call `sort(x)` and store the result (the sorted
vector) on a new object called `xs` ("x sorted").
```{r, fig.align = "center"}
# Sort x, store sorted vector on sx
xs <- sort(x)
xs
```

`xs` now contains the sorted version of `x` with the smallest
on the first element, the second largest on the second element
and so far and so on. **But** the length of `xs` is no longer
`r length(x)` (length of x) but `r length(xs)` (length of xs)!

**The optional argument `na.last`**

The reason is that `sort`, by default, uses `na.last = NA`.
If `na.last = NA` missing values will simply be **removed** from
the vector! In some situations we might need to keep the `NA`
values, and this can be controlled by using the additional
argument `na.last`.

Available options:

* `na.last = NA` (default): `NA`s are removed
* `na.last = TRUE`: `NA`s will be added to **the end** of the
     sorted vector ("place `NA` as last elements").
* `na.last = FALSE`: `NA`s will be added to **the beginning** of the
     sorted vector ("do not put `NA`s as last, but as first elements).

```{r, fig.align = "center"}
sort(x, na.last = NA) # default
sort(x, na.last = TRUE)
sort(x, na.last = FALSE)
```

As you can see `NA`s are once removed, once placed at the end of the
sorted vector, and once at the beginning of the sorted vector.

**The optional argument `decreasing`**

The other additional input argument `decreasing = TRUE` controls
whether the result is sorted in ascending (smallest to largest)
or descending (largest to smallest) order. For those using
SQL: this is equivalent to `"ORDER BY `field_name` ASC"` and
`"ORDER BY `field_name` DESC"`.

```{r, fig.align = "center"}
# Sort x descending
sort(x, decreasing = TRUE)
```


## Order of a vector

While `sort` _sorts_ our vector, `order` returns us the _order_ (index) of the
position of the elements in ascending/descending order.
From the _R_ documentation (`?order`/`help("order")`):

```
'order' returns a permutation which rearranges its first argument
into ascending or descending order, breaking ties by further
arguments. 'sort.list' is the same, using only one argument.
See the examples for how to use these functions to sort data
frames, etc.

order(..., na.last = TRUE, decreasing = FALSE,
      method = c("auto", "shell", "radix"))
```


Let us use the same vector `x` we used before, but call `order`
instead of `sort` this time:

```{r, fig.align = "center"}
x # Show x
which(is.na(x)) # Position of NA
# Order of vector x
order(x)
```


The result is an integer vector of the same length as the vector
`x`. By default, `order` returns an ascending order. Thus, the first
element returned by `order` gives us the **position of the smallest
value** in `x`, the second element the **second smallest**, and so far,
and so on. Here, the first element of `order(x)` (`order(x)[1L]`) is
`r order(x)[1L]` which simply tells us that **the smallest value in `x`
is on `x[`r order(x)[1L]``.

Remember that we have one missing value in `x` at position 8 (`x[8L]`)!
In contrast to `sort`, `order` does not remove missing values by default
but uses `na.last = TRUE`. Thus, we find the index `x` at the end of the
index vector returned by `order(x)`.

However, if needed we can also control the `na.last` behaviour as for the
`sort` function, except that the default is different:

```{r, fig.align = "center"}
order(x, na.last = NA)
order(x, na.last = TRUE) # default
order(x, na.last = FALSE)
```


Acending/decending works the same way, if we set `decreasing = TRUE`
_R_ will return the resverse order (index of the largest element first):

```{r, fig.align = "center"}
# Sort x descending
order(x, decreasing = TRUE)
```


## Sort to get order, order to sort

### Sort to get order

Sort and order do something similar but not the same.
To get a better understanding let's play arund a bit
using the following vector `x`:
```{r, fig.align = "center"}
set.seed(333)
x <- sample(1:100, 5)
x
sort(x)
order(x)
```


`x` is a `r class(x)` vector of length `r length(x)` containing
the values "`r paste(x, collapse = ", ")`". The question is:
**How could we use `sort` to get the order
of the elements in `x` without calling `order`)?**

Well, let's think about it for a second.
`sort` returns us the sorted vector of `x`.
Thus, we could first use `sort(x)` and store
the result on a new object `xs`. We know that
`order` returns us the position of the elements
in the original vector `x`. All we have to do
is to find out at which position in `x`
the smallest element is, and the smallest
element is the one stored on `xs[1L]`!
`xs[2L]` contains the second smallest element, ...

Thus, we could do the following:

* Sort `x` by calling `xs <- sort(x)`
* Create a new empty vector `idx` to store the position
   of the element `xs[i]` in `x`
* Create a for loop wich goes along all elements
   in `xs` (loop over `1:length(xs)`)

```{r, fig.align = "center"}
# Create new object 'xs',
# sorted version of x
xs <- sort(x, na.last = TRUE)
# Empty vector to store index
idx <- vector("integer", length(x))
```

```{r, fig.align = "center"}
# Loop over 'xs', find 'xs[i]' in x
for (i in seq_along(xs)) {
  idx[i] <- which(x == xs[i])
}
idx
```

Et voila, we get the very same as when calling `order` (compare
the result). **Warning:** will not work if `x` contains `NA`s.


### Order to sort

Another question is how to use `order` to sort a vector without
calling `sort`! This is more straight forward and can be used
to easily sort matrices and data frames.
Let's use the same vector `x` as before:

```{r, fig.align = "center"}
set.seed(333)
x <- sample(1:100, 5)
x
sort(x)
order(x)
```


As we already know `order` returns us the index of
the elements in `x` in an ascending order (by default).
Let's store the `order(x)` on a new object `xorder`.
Thus, `xorder[1L]` contains the index of the smallest
element in `x`, `xorder[2L]` the index of the second
smallest element in `x` and so far and so on.

All we have to do to bring `x` in an ascending order
is to pick the elements of `x` with the index sequence
of `order`. As `xorder` is an integer index vector
we can simply call `x[xorder]`:

```{r, fig.align = "center"}
# Get order of x
xorder <- order(x)
```

```{r, fig.align = "center"}
# Pick elements in 'x' in the
# order of 'xorder'
x[xorder]
```

... and done! `x[order]` (or, `x[order(x)]`  if you prefer
to write all in one go) returns us the very same result
as `sort(x)`.

## Sort a matrix

If we would like to sort a matrix we cannot simply use `sort`
anymore. The reason: a matrix is basically a vector. Thus,
`sort(x)` would return us **ALL** elements of the matrix
`x` in an ascending/descending order (to be more specific:
`sort(x)` of a matrix `x` returns a sorted **vector**).
Thus, to be able to sort a matrix by a specific column,
we have to do it slightly different.

What we have to do is to:

* Create an order sequence (index of the rows of a
    specific column in the matrix).
* Rearrange the matrix `x` given the oder sequence
    from the first step.

This will reorder/sort our matrix in a way that 
we will have an ascending/descending order of
the data of a **specific column** of our matrix.
But let's do a practical example ...

Let us start with creating a relatively
simple $3 \times 3$ integer matrix with
with three columns called `"A"`, `"B"`, and `"C"`
as follows:

```{r, fig.align = "center"}
# Demo matrix
x <- matrix(c(10,  8,  8,  3,
               8,  3,  1,  4,
              20, 33, 11, 31),
            ncol = 3)
colnames(x) <- c("A", "B", "C")
x
```


**Task 1:** assume that we would like to have the
matrix sorted such that we have ascending values
in the first column (column `"A"`) without mixing
up the data. For each row, the **data should be 
the same** as before, we **only** want to **re-order the rows!**

To do so we can again use `order` on column `"A"` by calling
`order(x[, "A"])` (and store it on a new object `idx`).
The result (`idx`) is an integer vector which tells us
in which row we have the **smallest value in column `"A"`,
and so far and so on.

We can now use **matrix subsetting by index** on our rows
to get the **rows of `x`** in the correct order.

```{r, fig.align = "center"}
idx <- order(x[, "A"])
idx
# Re-order matrix (store on
# 'xs' (x sorted).
xs <- x[idx, ]
xs
```


As you can see `xs` contains the sorted version with
increasing values in column `xs[, "A"]`, but the
data in each row are kept the same (the first row
of `xs` is simply what has been row `r order(x[, "A"])[1L]`
in our matrix `x`).

We can do the very same but sort by column `"B"`:
```{r, fig.align = "center"}
idx <- order(x[, "B"])
idx
# Re-order matrix (store on
# 'xs' (x sorted).
xs <- x[idx, ]
xs
```


Sometimes it is required to not only sort by one specific column,
but by first sort the matrix by one column, and then by another
one. This can be done by calling `order` with more than one input
vector. In this example I would like to sort the matrix `x`
by column `"A"` first, and then (if needed) by column `"B"`.
The procedure is the very same: we first create an index
vector `idx` with the row indices and then re-order the 
matrix `x` and store it on `xs`:

```{r, fig.align = "center"}
idx <- order(x[, "A"], x[, "B"])
idx
# Re-order
xs <- x[idx, ]
xs
```

As you can see the result is different than before
(check column `"B"`; second and third element).


## Sorting a data.frame

The very same as for a `matrix` can be used for a `data.frame`.
In this example I am loading the `faithful` data set from the
_R_ dataset package.

```{r, fig.align = "center"}
# Loading the old faithful
# data set:
data("faithful")
head(faithful, n = 3)
```


All I want to do is to sort the `data.frame`
by  `eruption` (eruption time in seconds) in an increasing
or ascending order. First, we create a new index vector
`idx` with the row indices ...

```{r, fig.align = "center"}
idx <- order(faithful$eruption)
head(idx)
```


... which we can use to re-order our `data.frame`.
As I don't need the unsorted version of the object `faithful`
anymore I simply overwrite it (`faithful <- faithful[idx,]`):

```{r, fig.align = "center"}
# Re-order (overwrite the
# old object 'faithful' here)
faithful <- faithful[idx, ]
# Head of sorted object
head(faithful)
```


As `head(faithful, n = 6)` returns us the first 6 entries
of our `data.frame` this is a very convenient way to get
the 6 elements with the smallest `eruption` times if needed
(`y <- head(faithful)`).

## Practical Exercise


The second practical example is about the good old titanic.
You can find a file called `"titanic_with_age.rda"` on OLAT
(_Downloads/Files_) which contains some information about the
passengers of the RMS Titanic. Try to do the following step/answer
the following questions:

1. Investigate the loaded object(s).
2. Write a `for` loop to get/print the class of each variable.
3. Do we have missing values? Do we have wrong/suspicious values?
4. For numeric columns: calculate summary statistics (e.g.,
            mean, minimum, maximum, range, ...).
5. Sort the object by age (decreasing, oldest first).
6. Create a new factor variable `age_group` with `"child"` for passengers $\le
    14$ and `"adult"` for all others.
7. Create a subset for "females", one for "males" using `subset`.
8. Use `subset to create a new object only containing the variables `gender`
    and `alive`.
9. Was it more likely to survive as a female? Count number of males/females and
    (given gender) how may of them survived/died. Advanced: use `table` and
    `prop.table`

